Morse Theory

### 2020-11-29 paper.tex
dsads

\textbf{Orientable Manifolds} \\
Our goal is to express number of loops in a Reeb graph by # of critical points. Let $n_i$ be the number of nodes with degree $i$. Orientable manifolds have 0 $n_2$ because every point is either minimum, maximum or saddle.\\
Therefore the number of arcs is number of connections each point has, divided by two (because we count each arc twice) is $e = (n_1+3*n_3)/2$ \\ 
number of loops is $1+e-(n_1+n_3)
% % end M2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % begin V2
% % end V2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{9}
     \bibitem{s}
         {\sc }
         `` https://core.ac.uk/download/pdf/20670562.pdf
\end{thebibliography}

@@

\textbf{Non-orientable 2-manifolds without boundary}\\

We will consider in this section Reeb graphs of Morse functions over non-orientable 2-manifolds with boundary. In particular, number of nodes, contours and loops in such graphs. \\
All connected non-orientable 2-manifolds without boundary could be expressed as connected sum of  $g$ copies of projective planes $N=P\# P...\# P$, where $g$ coincides with the genus of the surface.\\
For the further proofs we have to recall that Euler characteristic of orientable surfaces can be calculated as $\chi=2-2g$ and for non-orientable it is $\chi=2-g$. Also we have to notice that we can make a 2-sheeted cover of a non-orientable 2-manifold, where each sheet is orientable. In future we will call this $doubling$. \\
Let's consider the process of doubling. We obtain 2-sheeted cover of $N$ (non-orientable 2-manifold) by doubling each point $x\in N$ to two points $x'$ and $x''$. We can consider them as a points lying in the neighborhood of $x$ on two opposite sides of $N$. Union of such points $x'$ and $x''$ will be space that is connected and orientable 2-manifold $M$. Due to the doubling Euler characteristic of the surface will be doubled, $\chi_M=2\chi_N$. \\
Now we have to calculate genus of $M$. Since it is orientable:\\ $g_M=\frac{2-\chi_M}{2}=1-\frac{\chi_M}{2}=1-\chi_N=1-(2-g_N-1)=g_N-1$. \\
Therefore, by doubling we decrease genus by 1. \\
In the table on the fig. \ref{fig:Table} one can find Euler characteristicand genus number of basic non-orianatble manifolds. Note that $K$ is a Klein bottle, $T$ represents torrus,$S$ sphere and $P$ is a projective plane. 
\begin{figure}[h!]
\center{\includegraphics[scale=0.9]{char_table.png}}
\caption{Euler characteristic and genus of some non-orientable manifolds and their doublings}
\label{fig:Table}
\end{figure}
On the table fig. \ref{fig:Table} one can observe that relation of Euler characteristic and genus number of non-orientable manifolds and their doublings coincides with the dependence introduced above. \\
Now we can move on to describing Reeb graphs of non-orientable 2-manifolds without boundary. As an example let's consider Klein bottle, its Reeb graph, and Reeb graph of its doubling surface on fig.\ref{fig:Klein}. \\
\begin{figure}[h!]
\center{\includegraphics[scale=0.45]{Klein_bottle.png}}
\caption{Reeb graph of Klein bottle and its doubling}
\label{fig:Klein}
\end{figure}
As we see on the fig. \ref{fig:Klein} Reeb graph of the surface drastically changes after doubling, especially number of nodes and contours.  Let's learn how do nodes of each degree change under this doubling. \\
1-degree nodes are maxima or minima, after doubling number of these nodes is doubled and also number of contours is doubled as well. \\
2-degree nodes exist in Reeb graphs of non-orientable surfaces, despite the fact that in orientable surfaces they could be just collapsed. It is transformed into 4-degree node after doubling. The reason is that the level set containing a critical point of index 1 is a figure-8 which may contain an orientation-reversing cycle. \\
3-degree nodes are saddle points that just transfer into two such nodes in its doubling. \\
One could observe the above on the fig. \ref{fig:Nodes}.  
\begin{figure}[h!]
\center{\includegraphics[scale=0.3]{Nodes.png}}
\caption{Correspondence between nodes in $N$ and its doubling $M$}
\label{fig:Nodes}
\end{figure}
Let $f: N \rightarrow R$ is a Morse function that induces function $f_0: M \rightarrow R$. $f_0$ is not a Morse function, since all critical points of $f$ correspond to two critical points of $f_0$, so they are not distinct. However, critical points of $f_0$ are still non-degenerate, that gives as an opportunity to consider $R(f_0)$ (Reeb graph of $f_0$) as previously. \\
Let $e$ be number of arcs in $R(f)$, $n_{1,3}$ is number of 1-degree and 3-degree nodes of $R(f)$, and $n_2$ is number 2-degree nodes. As seen previously, number of 1,3-degree nodes and their contours are doubled in $M$. However, number of 2-degree nodes is invariant under doubling, corresponding contours are doubled. \\
Therefore, $R(f_0)$ has $2e$ arcs and $2n_{1,3}+n_2$ nodes. Known fact that number of loops in a graph is $\beta_1 (G)=1-+e-n$, where $e$ is number of edges(arcs) and $n$ is number of nodes. \\
After substitution we get:\\
$1-\beta_1(R(f_0)=2n_{1,2}+n_2-2e=2-2+2n_{1,3}+2n_2+2e-n_2=2-2(1-n_{1,3}-n_2+e)-n_2=2-2\beta_1(R(f))-n_2$. \\
Also we know that number of loops of $N$ is its genus and genus of $M$ (number of loops) is one less that one of $N$. Therefore, we can make a substitution $\beta_1(R(f_0))=g_N-1$. \\
From the above results we get: \\
$\beta_1(R(f))=\frac{g_N-n_2}{2}$.\\
Since number of 2-degree nodes $n_2\geq0 $, we may obtain the following result:\\
\textbf{Lamma ?}\\
The Reeb graph of a Morse function over connected non-orientable 2-manifolds of genus $g$ with boundary is at most $\frac{g}{2}$.\\

>>>>

\textbf{Orientable Manifolds} \\
Our goal is to express number of loops in a Reeb graph by # of critical points. Let $n_i$ be the number of nodes with degree $i$. Orientable manifolds have 0 $n_2$ because every point is either minimum, maximum or saddle.\\
Therefore the number of arcs is number of connections each point has, divided by two (because we count each arc twice) is $e = (n_1+3n_3)/2$ \\ 
number of loops is $1+e-(n_1+n_3)$
\begin{definition}
    Let P be a polyhedron with V vertices, E edges, and F faces. Then we define the Euler characteristic to be 
    \begin{equation}
        \chi(P) = V - E + F  
    \end{equation}
\end{definition} \\
More generally, Euler characteristic is defined as alternating sum of Betti numbers. Last Morse inequality allow us to define $\chi$ as a set of inequalities for alternating sums of Betti numbers in terms of a corresponding alternating sum of the number of critical point $n_i$ of a Morse function of a given index.
\begin{lemma}
    The Reeb graph of a Morse function on a connected, orientable 2-manifold of genus g has g loops.
    \begin{proof}
        We first suppose that Reeb graph has no loops. It has $n_1 = n_3+2$. And if we write $c_i$ for the number of critical points of index $i$ we have $n_i = c_0+c_2$ and $n_3 = c_1$. we have $\chi = c_1 - c_1 +c_2 = n_1 - n_3 = 2.$  2 is the Euler characteristic of a sphere, which has no genus. \\ 
        Now suppose that Reeb graph has at least one loop. We can apply following two functions because they preserve homotopy type and the number of loops: collapse degree 1 nodes and merge arcs across degree 2 nodes which get eliminated in the process. For example consider the following graph:
    \begin{center}
    \includegraphics[width=0.4\textwidth]{homotopic_transformation.png}
    \end{center}
    Let $m_3$ be the number of remaining degree 3 nodes and note that it is even because $3m_3$ is twice the number of remaining arcs. Using the Euler-Poincare Theorem for graphs we get $\chi = m_3 - 3m_3/2 = \beta_0 - \beta_1$. The graph is connected which implies that the number of loops is $\beta = m_3/2+1$. We have $c_1$ degree 3 nodes in the original Reeb graph and for each minimum and maximum we collapse one degree 1 node removing a 3 degree node in the process. Last strong Morse inequality tells us that $m_3 = c_1 - (c_0 + c_2) = -\chi = 2g - 2$. The number of loops is therefore $ \beta_1 = (2g-2)/2+1=g$
    \end{proof}
\end{lemma}
% % end M2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % begin V2
% % end V2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{9}
     \bibitem{s}
         {\sc }
         `` https://core.ac.uk/download/pdf/20670562.pdf
\end{thebibliography}

@@
